Bacon: A GPU Programming Language With Just in Time Specialization

### 2011-05-30 doc/Paper/Paper.tex
Bacon paper: Now with less fail.

Benchmarks are provided for matrix multiplication comparing a Bacon
implementation to versions using OpenCL directly. Speedups are
demonstrated both for naive implementations and when comparing a naive
Bacon implementation to a hand-vectorized OpenCL kernel. This latter
result demonstrates the benefit of the total loop unrolling enabled by
just-in-time specialization. Additionally, a Bacon implementation of a
more complex computation, stereo disparity, is considered.

@@

Bacon is not the first attempt to provide an improved programmer
experience for OpenCL.

@@

    float sum = 0.0;
    assert(aa.cols == bb.rows, 
        "Matrices must have compatible dimensions.");
    for (int kk = 0; kk < aa.cols; ++kk) {
        sum += aa[$row, kk] * bb[kk, $col];
    }
    cc[$y, $x] = sum;
    return cc;
}
\end{verbatim}
\caption{Naive Matrix Multiplication in Bacon C}\label{mmk}
\end{listing}

@@

Parametrized types for 1D, 2D and 3D arrays are provided
natively. The types are parametrized using C++-style angle bracket
syntax. Both declarations and element access use a comma separated
list of dimensions in square brackets. The dimensions of these arrays
can be accessed using struct-style dot notation.

@@

This specialization, in addition to providing performance benefits,
allows for the simulation of variable sized arrays in thread-local
memory as long as the array size depends only on {\tt const} variables
and array dimensions. The blocked matrix multiply kernel in
Listing~\ref{bmmk} gives an example of this feature.

@@

The Bacon system consists of two pieces: the front-end compiler and the
runtime library. The front-end library parses the Bacon C source and
outputs a C++ wrapper and an intermediate form of the parsed code. The
runtime library is called from the generated wrapper to load the
intermediate form code, generate specialized OpenCL C code when a
kernel is called, and run that code on the GPU using the OpenCL
runtime.

The system is built using Perl and C++. The front-end compiler parses
the source code using Parse::Yapp\cite{parse::yapp}, a yacc-compatible
parser generator for Perl. This constructs abstract syntax tree, which
is then walked to generate the C++ wrapper. The abstract syntax tree
(AST) is then serialized for later loading by the runtime library.

@@

The two optimizations that are performed by the Bacon library are
constant propagation and loop unrolling. Constant propagation is used
to determine the exact set and values of the specialized variables,
building a mapping {\tt const} variables to their values so that
references to these variables can be replaced with integer literals in
the generated OpenCL C code. Loop unrolling is then performed on any
loops that can be statically analized.

@@

In order to evaluate the performance of the Bacon system, we compare
the run time of matrix multiplication kernels in both Bacon C and
OpenCL C. These results are summarized in Table~\ref{mm1}, which shows
that Bacon is able to provide measurable performance improvements over
similar programs written directly in OpenCL C.

@@

\begin{description}
  \item[Bacon - Naive] Shown in Listing~\ref{mmk}.
  \item[OpenCL - Naive] An equivalent OpenCL C implementation.
  \item[Bacon - Blocked] Shown in listing~\ref{bmmk}.
  \item[OpenCL - Hand Vectorized] Hand unrolled using vector types to
    compute 4x4 blocks. Based on a sample from the AMD OpenCL SDK.
\end{description}

Timings were taken multiplying 4096x4096 matrices. Each test was
performed five times and the average result was taken. The times were
very consistent; most tests had a coefficient of variation under one
percent. The speedups over the naive OpenCL implementation are shown
in Figure~\ref{unroll}.

@@

These measurements clearly show that value specialization provides
significant speedups over a non-specialized OpenCL kernel. This result
was expected; providing constant values at compile time allows the
OpenCL C compiler to do extensive and agressive constant
propagantion-based optimizations.

@@

>>>>

Benchmarks are provided for matrix multiplication comparing a Bacon
implementation to versions using OpenCL directly. Speedups are
demonstrated both for naive implementations and when comparing a Bacon
implementation of generalized block decomposed matrix multiplication
to a hand-vectorized OpenCL kernel. This latter result demonstrates
the benefit of the total loop unrolling enabled by just-in-time
specialization. Additionally, a Bacon implementation of a more complex
computation, stereo disparity, is considered.

@@

Bacon is not the first attempt to provide an improved programmer
experience for OpenCL. Rick Webber has developed an improved C++ API
called clUtil\cite{webber:2011}. Bindings for other languages, like
the JOCL\cite{hutter:2011} binding for Java, provide APIs at a variety
of levels of abstraction.

Another very interesting approach allows the programmer to describe
the computation to be executed on the GPU in the same language as the
rest of the program. An OpenCL kernel is then generated
automatically. This has the potential benefit of increased expressive
power and programmer familiarity at the cost of increased complexity
and necessarily leaky abstraction when the host language contains
features that can't be cleanly transformed into GPU code. Examples of
this approach include CLyther\cite{rossross:2011} for Python and
ScalaCL\cite{chafik:2011} for the functional language Scala.

@@

    float sum = 0.0;
    assert(aa.cols == bb.rows, 
        "Matrices must have compatible dimensions.");
    for (int kk = 0; kk < aa.cols; ++kk) {
        sum += aa[$row, kk] * bb[kk, $col];
    }
    cc[$row, $col] = sum;
    return cc;
}
\end{verbatim}
\caption{Naive Matrix Multiplication in Bacon C}\label{mmk}
\end{listing}

@@

Parametrized types for 1D, 2D, and 3D arrays are provided
natively. The types are parametrized using C++-style angle bracket
syntax. Both declarations and element access use a comma separated
list of dimensions in square brackets. The dimensions of these arrays
can be accessed using struct-style dot notation.

@@

This specialization, in addition to providing performance benefits,
allows for the simulation of variable sized arrays in thread-local
memory as long as the array size depends only on {\tt const} variables
and array dimensions. This works around one of the main constraints
imposed by OpenCL. The blocked matrix multiply kernel in
Listing~\ref{bmmk} gives an example of this feature.

@@

The Bacon system consists of two pieces: the compiler and the runtime
library. The compiler parses the Bacon C source and outputs a C++
wrapper and the serialized abstract syntax tree (AST). The runtime
library is called from the generated wrapper to load the intermediate
form code, generate specialized OpenCL C code when a kernel is called,
and run that code on the GPU using the OpenCL runtime.

The system is built using Perl and C++. The front-end compiler parses
the source code using Parse::Yapp\cite{parse::yapp}, a yacc-compatible
parser generator for Perl. This constructs the abstract syntax tree,
which is then walked to generate the C++ wrapper.

@@

The two optimizations that are performed by the Bacon library are
constant propagation and loop unrolling. Constant propagation
calculates the values of all the variables that have been marked as
\tt{const} by the programmer. If the value of any of these variables
cannot be computed from the specialized arguments to the kernel, the
Bacon runtime will throw an exception.  This information is used to
construct a symbol table, and references to these variables are
replaced with their constant integer values in the generated OpenCL
code. Loop unrolling is then performed on any loops that can be
statically analyzed.

@@

In order to evaluate the performance of the Bacon system, we compare
the run time of matrix multiplication kernels written in both Bacon C
and OpenCL C. These results are summarized in Table~\ref{mm1}, which
shows that Bacon is able to provide measurable performance
improvements over similar programs written directly in OpenCL C.

@@

\begin{description}
  \item[Bacon - Naive] Shown in Listing~\ref{mmk}.
  \item[OpenCL - Naive] An equivalent OpenCL C implementation.
  \item[Bacon - Blocked] Shown in Listing~\ref{bmmk}.
  \item[OpenCL - Hand Vectorized] Hand unrolled using vector types to
    compute 4x4 blocks. Based on a sample from the AMD OpenCL SDK.
\end{description}

The execution time of these four kernels was tested on randomly
generated 4096x4096 matrices. Each test was performed five times and
the average result was taken. The times were very consistent; most
tests had a coefficient of variation under one percent. The speedups
over the naive OpenCL implementation are shown in Figure~\ref{unroll}.

@@

These measurements clearly show that value specialization provides
significant speedups over a non-specialized OpenCL kernel. This result
can be explained by the fact taht providing constant values at compile
time allows the OpenCL C compiler to do extensive constant
propagation-based optimizations.

@@

\section{Stereo Disparity in Bacon}

In order to demonstrate that Bacon is suitable for complex
applications we consider an implementation of stereo disparity
matching. This is an application that is well suited to GPU
implementation due to being computationally expensive and naturally
parallel. The stereo disparity computation is significantly more
complex than matrix multiplication, requiring multiple steps such that
the entire process can not easily be implemented as a single kernel.

The stereo disparity matching problem is to generate a disparity image
given a pair of images from a stereo camera rig. Each pixel ($y,x$) in
the disparity image has a value ($v$) specifying a pixel offset that
matches the pixel at ($y,x$) in the left image to the pixel at
($y,x+v$) in the right image. These matched pixels correspond to the
same physical point in scene viewed by the cameras. This disparity
image can then be used to compute depth information and create a 3D
point cloud describing the scene, which is useful in a variety of
practical applications like robotics.

The stereo disparity method implemented used the Census
transform\cite{zabih:1994} both with simple local window matching and
with a partial implementation of the semi-global matching (SGM) technique
described by Hirschmuller\cite{hirschmuller:2005}. Testing during
development was done on sample image pairs from the Middlebury 2005
Stereo Dataset\cite{hirschmuller:2005}. Both a CPU implementation and
a Bacon implementation were written.

Although a complete implementation of the SGM algorithm was not
completed and neither Bacon implementation has been properly
optimized, the results are promising. The simple local window matcher
in Bacon produces the same output as the CPU version. The partial SGM
implementation produces a more accurate disparity image but takes
longer to run.

Including specialization time, the Bacon version of the local window
matcher takes almost as much time as the CPU version, but on
subsequent calls it takes less than 0.2 seconds which is more than a
10x speedup. This provides a good example of the practicality of the
value specialization mechanism; for processing video an extra
computation cost on the first frame will be quickly amortized over
subsequent frames.
